---
title: "Case Study: Bike Sharing System"
author: "Winston Koh"
format: 
  revealjs:
    theme: simple
    embed-resources: true
    slide-number: true
    preview-links: true
---

```{r setup}
#| include: false

library(tidyverse)
library(arrow)
library(janitor)
library(leaflet)
options(dplyr.summarise.inform = FALSE)

# Combined and cleaned data
data2 <- read_parquet("combined2024-divvy-tripdata_clean.parquet")

# Number of trips by end location
end_trips_n <- data2 |> 
  group_by(member_casual, end_station_id, end_station_name, end_lat, end_lng) |> 
  summarise(n = n())

# The same station can have multiple different coordinates
# This table gives one coordinate for each station using the most frequently appearing coordinate
end_locations <- data2 |> 
  group_by(end_station_id, end_station_name, end_lat, end_lng) |> 
  summarise(n = n()) |> # Number of trips for each combination of id, name, lat, lng
  ungroup() |> 
  group_by(end_station_id) |> 
  slice_max(n) |> # Most frequent lat and lng for each id
  ungroup() |> 
  select(-n)
```

## Background

- Cyclistic is a fictional bike-share company in Chicago (based on the real-life [Divvy](https://divvybikes.com/about) bike-share program).
- Casual riders are customers who purchase single-ride or full-day passes.
- Members are customers who purchase annual memberships.

## Problem

- Annual members are much more profitable than casual riders.
- Maximizing the number of annual members will be key to future growth.
- We want to design a marketing strategy to convert casual riders into members.

## Questions of Interest

1. How do annual members and casual riders use Cyclistic bikes differently?
2. Why would casual riders buy Cyclistic annual memberships?
3. How can Cyclistic use digital media to influence casual riders to become members?

## Data

We have [publicly available](https://divvybikes.com/system-data), anonymized data from Divvy for the year 2024. Each trip has the following data:

- Rider type (Casual or Member)
- Trip start date + time and end date + time
- Trip start location and end location
- Ride type (Electric bike, Classic bike, or Electric scooter)

## Findings

Most rides are taken by members, and member trips are usually shorter than casual rider trips.

:::: {.columns}

::: {.column width="45%"}
```{r customer type pie chart}
#| echo: false
#| fig-width: 6
#| fig-height: 6

data2 |> 
  group_by(member_casual) |> 
  summarise(n = n()) |> 
  mutate(prop = n / nrow(data2)) |> 
  ggplot(aes(x = "", y = n, fill = member_casual)) +
  geom_bar(stat = "identity") +
  coord_polar("y", start=0) +
  geom_text(
    aes(label = paste0(round(prop, 2) * 100, "%")), 
    position = position_stack(vjust=0.5),
    size = 6
  ) +
  labs(x = NULL, y = NULL, fill = NULL,
       title = "Proportion of Rides by Customer Type",
       subtitle = "Total # of Rides = 5,859,845") +
  theme_void() +
  theme(plot.title = element_text(hjust = 0.5, size = 18),
        plot.subtitle = element_text(hjust = 0.5, size = 14),
        legend.text = element_text(size = 14))
```
:::

::: {.column width="10%"}
 
:::

::: {.column width="45%"}
```{r member casual duration}
#| echo: false
#| fig-width: 6
#| fig-height: 6

data2 |> 
  group_by(member_casual) |> 
  summarise(median_duration = median(ride_length)) |> 
  ggplot(aes(x = member_casual, y = as.numeric(median_duration)/60, fill = member_casual)) +
  geom_bar(stat = "identity") +
  theme_bw() +
  labs(x = "Customer Type",
       y = "Duration (min)",
       title = "Median Ride Length") +
  theme(plot.title = element_text(hjust = 0.5, size = 18),
        axis.text = element_text(size = 12),
        axis.title = element_text(size = 14),
        legend.position = "none")
```
:::

::::

## Trips by Hour

Both member and casual rider trips peak in the afternoon, though members have a small peak in the morning.

```{r}
#| echo: false

data2 |> 
  group_by(member_casual, start_hour) |> 
  summarise(n = n()) |> 
  ggplot(aes(x = start_hour, y = n/1000)) +
  geom_col() +
  facet_grid(member_casual~.) +
  theme_bw() +
  labs(title = "Number of Trips by Hour",
       x = "Hour",
       y = "# of Trips (thousands)") +
  theme(legend.position = "none",
        plot.title = element_text(hjust = 0.5, size = 16),
        axis.ticks = element_blank())
```

## Trips by Day of Week

Members take more trips during weekdays, while casual riders take more trips during weekends.

```{r trips weekday bar chart}
#| echo: false

data2 |> 
  group_by(member_casual, day_of_week) |> 
  summarise(n = n()) |> 
  ggplot(aes(x = day_of_week, y = n/1000)) +
  geom_col() +
  facet_wrap(~member_casual) +
  theme_bw() +
  labs(title = "Number of Trips by Day of Week",
       x = "Day of Week",
       y = "# of Trips (thousands)") +
  theme(legend.position = "none",
        plot.title = element_text(hjust = 0.5, size = 16),
        axis.ticks = element_blank())
```

## 

Member trip durations are relatively constant throughout the week, while casual rider trip durations are longer on the weekends.

```{r duration weekday bar chart}
#| echo: false

data2 |> 
  group_by(member_casual, day_of_week) |> 
  summarise(median_duration = median(ride_length)) |> 
  ggplot(aes(x = day_of_week, y = as.numeric(median_duration)/60)) +
  geom_bar(stat = "identity") +
  facet_wrap(~member_casual) +
  theme_bw() +
  labs(x = "Day of Week",
       y = "Duration (min)",
       title = "Median Ride Length by Day of Week") +
  theme(plot.title = element_text(hjust = 0.5, size = 16))
```

## Trips by Month

Trips are more frequent in the warmer months for both members and casual riders.

```{r trips month bar chart}
#| echo: false

data2 |> 
  group_by(member_casual, month) |> 
  summarise(n = n()) |> 
  ggplot(aes(x = month, y = n/1000)) +
  geom_col() +
  facet_wrap(~member_casual) +
  theme_bw() +
  labs(title = "Number of Trips by Month",
       x = "Month",
       y = "# of Trips (thousands)") +
  theme(legend.position = "none",
        plot.title = element_text(hjust = 0.5, size = 16),
        axis.ticks = element_blank())
```

## 

Casual rider trips are much longer in the warmer months compared to the colder months. This difference is much smaller for members.

```{r duration month bar chart}
#| echo: false

data2 |> 
  group_by(member_casual, month) |> 
  summarise(median_duration = median(ride_length)) |> 
  ggplot(aes(x = month, y = as.numeric(median_duration)/60)) +
  geom_bar(stat = "identity") +
  facet_wrap(~member_casual) +
  theme_bw() +
  labs(x = "Month",
       y = "Duration (min)",
       title = "Median Ride Length by Month") +
  theme(plot.title = element_text(hjust = 0.5, size = 16))
```


## Trips by Destination

The 20 most frequent destinations of casual rider trips are recreational areas by the waterfront.

```{r casual end dest}
#| echo: false

casual_dest <- end_trips_n |> 
  filter(member_casual == "Casual") |> 
  group_by(end_lat, end_lng) |> 
  summarise(rides = sum(n)) |> 
  arrange(desc(rides)) |> 
  left_join(end_locations, c("end_lat", "end_lng")) # Add available station names

casual_dest |> 
  head(20) |> 
  leaflet() |> 
	addTiles() |>
	addCircles(
	  lng = ~end_lng,
	  lat = ~end_lat,
	  weight = 5,
	  radius = ~sqrt(rides),
	  popup = ~end_station_name
	)
```

## 

The 20 most frequent destinations of member trips include the downtown area, transit centers, and The University of Chicago. 

```{r member end dest}
#| echo: false

member_dest <- end_trips_n |> 
  filter(member_casual == "Member") |> 
  group_by(end_lat, end_lng) |> 
  summarise(rides = sum(n)) |> 
  arrange(desc(rides)) |> 
  left_join(end_locations, c("end_lat", "end_lng"))


member_dest |> 
  head(20) |> 
  leaflet() |> 
	addTiles() |>
	addCircles(
	  lng = ~end_lng,
	  lat = ~end_lat,
	  weight = 5,
	  radius = ~sqrt(rides),
	  popup = ~end_station_name
	)
```

## Summary

Data suggests that casual riders tend to take trips mainly for leisure while members tend to take trips for daily commuting.

- Casual riders occur more frequently and tend to be longer during weekends and warmer months. 
The most common destinations are recreational areas on the waterfront. 
- Member trips occur more frequently during weekdays, and have a spike in the morning hours. Trips are shorter and more consistent in duration compared to casual rider trips. 
The most common destinations are in the downtown area. 

## Recommendations

- Convince casual riders to daily bike commuters by emphasizing the benefits of commuting by bike instead of by other forms of transportation.
  - e.g., Widespread availability, cost, speed, healthy physical activity
- Create promotional deals for members for leisure activities and destinations 
  - e.g., The aquarium, museums, restaurants

