---
title: "Case Study: Bike Sharing System"
author: "Winston Koh"
format: 
  revealjs:
    theme: simple
    embed-resources: true
    slide-number: true
    preview-links: true
---

```{r setup}
#| include: false

library(tidyverse)
library(arrow)
library(janitor)
library(leaflet)
options(dplyr.summarise.inform = FALSE)

# Combined and cleaned data
data2 <- read_parquet("combined2024-divvy-tripdata_clean.parquet")

# Number of trips by end location
end <- data2 |> 
  group_by(member_casual, end_station_id, end_station_name, end_lat, end_lng) |> 
  summarise(n = n())

# The same station can have multiple different coordinates
# This table gives one coordinate for each station using the most frequently appearing coordinate
end_locations <- data2 |> 
  group_by(end_station_id, end_station_name, end_lat, end_lng) |> 
  summarise(n = n()) |> 
  ungroup() |> 
  group_by(end_station_id) |> 
  slice_max(n) |> 
  ungroup() |> 
  drop_na() |> 
  select(-n)
```

## Background

- Cyclistic is a fictional bike-share company in Chicago (based on the real-life [Divvy](https://divvybikes.com/about) bike-share program).
- Casual riders are customers who purchase single-ride or full-day passes.
- Members are customers who purchase annual memberships.

## Problem

- Annual members are much more profitable than casual riders.
- Maximizing the number of annual members will be key to future growth.
- We want to design a marketing strategy to convert casual riders into members.

## Questions of Interest

1. How do annual members and casual riders use Cyclistic bikes differently?
2. Why would casual riders buy Cyclistic annual memberships?
3. How can Cyclistic use digital media to influence casual riders to become members?

## Data

We have data on bike trips from the year 2024 with the following data for each trip:

- Customer type (Casual or Member)
- Trip start date + time and end date + time
- Trip start location and end location
- Ride type (Electric bike, Classic bike, or Electric scooter)

## Data Considerations

- The data originates from [Divvy's](https://divvybikes.com/system-data) publicly available data.
- Each trip is anonymized.
- Trips that were below 60 seconds in length are excluded.
  - These are potentially false starts or users trying to re-dock a bike to ensure it was secure.

## Findings

Most rides are taken by members, and member trips are usually shorter than casual rider trips.

:::: {.columns}

::: {.column width="45%"}
```{r customer type pie chart}
#| echo: false
#| fig-width: 6
#| fig-height: 6

data2 |> 
  group_by(member_casual) |> 
  summarise(n = n()) |> 
  mutate(prop = n / nrow(data2)) |> 
  ggplot(aes(x = "", y = n, fill = member_casual)) +
  geom_bar(stat = "identity") +
  coord_polar("y", start=0) +
  geom_text(
    aes(label = paste0(round(prop, 2) * 100, "%")), 
    position = position_stack(vjust=0.5),
    size = 6
  ) +
  labs(x = NULL, y = NULL, fill = NULL,
       title = "Proportion of Rides by Customer Type",
       subtitle = "Total # of Rides = 5,859,845") +
  theme_void() +
  theme(plot.title = element_text(hjust = 0.5, size = 18),
        plot.subtitle = element_text(hjust = 0.5, size = 14),
        legend.text = element_text(size = 14))
```
:::

::: {.column width="10%"}
 
:::

::: {.column width="45%"}
```{r member casual duration}
#| echo: false
#| fig-width: 6
#| fig-height: 6

data2 |> 
  group_by(member_casual) |> 
  summarise(median_duration = median(ride_length)) |> 
  ggplot(aes(x = member_casual, y = as.numeric(median_duration)/60, fill = member_casual)) +
  geom_bar(stat = "identity") +
  theme_bw() +
  labs(x = "Customer Type",
       y = "Duration (min)",
       title = "Median Ride Length") +
  theme(plot.title = element_text(hjust = 0.5, size = 18),
        axis.text = element_text(size = 12),
        axis.title = element_text(size = 14),
        legend.position = "none")
```
:::

::::

## Trips by Hour

Both member and casual rider trips peak in the afternoon, though members have a small peak in the morning.

```{r}
#| echo: false

data2 |> 
  group_by(member_casual, start_hour) |> 
  summarise(n = n()) |> 
  ggplot(aes(x = start_hour, y = n/1000)) +
  geom_col() +
  facet_grid(member_casual~.) +
  theme_bw() +
  labs(title = "Number of Trips by Hour",
       x = "Hour",
       y = "# of Trips (thousands)") +
  theme(legend.position = "none",
        plot.title = element_text(hjust = 0.5, size = 16),
        axis.ticks = element_blank())
```

## Trips by Day of Week

Members take more trips during weekdays, while casual riders take more trips during weekends.

```{r trips weekday bar chart}
#| echo: false

data2 |> 
  group_by(member_casual, day_of_week) |> 
  summarise(n = n()) |> 
  ggplot(aes(x = day_of_week, y = n/1000)) +
  geom_col() +
  facet_wrap(~member_casual) +
  theme_bw() +
  labs(title = "Number of Trips by Day of Week",
       x = "Day of Week",
       y = "# of Trips (thousands)") +
  theme(legend.position = "none",
        plot.title = element_text(hjust = 0.5, size = 16),
        axis.ticks = element_blank())
```

## 

Member trip durations are relatively constant throughout the week, while casual rider trip durations are longer on the weekends.

```{r duration weekday bar chart}
#| echo: false

data2 |> 
  group_by(member_casual, day_of_week) |> 
  summarise(median_duration = median(ride_length)) |> 
  ggplot(aes(x = day_of_week, y = as.numeric(median_duration)/60)) +
  geom_bar(stat = "identity") +
  facet_wrap(~member_casual) +
  theme_bw() +
  labs(x = "Day of Week",
       y = "Duration (min)",
       title = "Median Ride Length by Day of Week") +
  theme(plot.title = element_text(hjust = 0.5, size = 16))
```

## Trips by Month

Trips are more frequent in the warmer months for both members and casual riders.

```{r trips month bar chart}
#| echo: false

data2 |> 
  group_by(member_casual, month) |> 
  summarise(n = n()) |> 
  ggplot(aes(x = month, y = n/1000)) +
  geom_col() +
  facet_wrap(~member_casual) +
  theme_bw() +
  labs(title = "Number of Trips by Month",
       x = "Month",
       y = "# of Trips (thousands)") +
  theme(legend.position = "none",
        plot.title = element_text(hjust = 0.5, size = 16),
        axis.ticks = element_blank())
```

## 

During the warmer months, members trips are slightly longer, while casual rider trips are much longer.

```{r duration month bar chart}
#| echo: false

data2 |> 
  group_by(member_casual, month) |> 
  summarise(median_duration = median(ride_length)) |> 
  ggplot(aes(x = month, y = as.numeric(median_duration)/60)) +
  geom_bar(stat = "identity") +
  facet_wrap(~member_casual) +
  theme_bw() +
  labs(x = "Month",
       y = "Duration (min)",
       title = "Median Ride Length by Month") +
  theme(plot.title = element_text(hjust = 0.5, size = 16))
```


## Trips by Destination

The 20 most frequent destinations of casual rider trips are recreational areas by the waterfront.

```{r casual end dest}
#| echo: false

casual_dest <- end |> 
  filter(member_casual == "Casual") |> 
  group_by(end_lat, end_lng) |> 
  summarise(rides = sum(n)) |> 
  arrange(desc(rides))

casual_dest |> 
  head(20) |> 
  left_join(end_locations, c("end_lat", "end_lng")) |> 
  leaflet() |> 
	addTiles() |>
	addCircles(
	  lng = ~end_lng,
	  lat = ~end_lat,
	  weight = 5,
	  radius = ~sqrt(rides),
	  popup = ~end_station_name
	)
```

## 

The 20 most frequent destinations of member trips include the downtown area, transit centers, and The University of Chicago. 

```{r member end dest}
#| echo: false

member_dest <- end |> 
  filter(member_casual == "Member") |> 
  group_by(end_lat, end_lng) |> 
  summarise(rides = sum(n)) |> 
  arrange(desc(rides))

member_dest |> 
  head(20) |> 
  left_join(end_locations, c("end_lat", "end_lng")) |> 
  leaflet() |> 
	addTiles() |>
	addCircles(
	  lng = ~end_lng,
	  lat = ~end_lat,
	  weight = 5,
	  radius = ~sqrt(rides),
	  popup = ~end_station_name
	)
```

## Summary

- Casual riders tend to take longer bike trips (especially on weekends and during the warmer months) to recreational areas on the waterfront. 
- Members tend to take shorter, more consistently long trips to the downtown area.
- This suggests that casual riders tend to take trips for leisure while members tend to take trips for commuting to work/school.

## Recommendations

- Conduct a survey to confirm this hypothesis
- Try to convert casual riders to use our bike-sharing service for commuting and regular use.
- Use digital media to emphasize the benefits of commuting by bike over other forms of transit.

